Sketching for Density Estimation

### 2018-12-17 komlos.tex
changing definitions

\section{Komlos Complexity}
We begin by formally defining the Komols complexity and by giving some general results about for problems with low Komols complexity. The Komols complexity of a sequence set is defined similarly to it Rademacher complexity.
\begin{definition}
Let $A \subset [0,1]^m$ and $\sigma \in \{-1,1\}^m$ uniformly. 
The Rademacher complexity of $\F$ is 
\[
R_m(\F) =  \frac{1}{m}\E_{\sigma} \sup_{a \in A}  \sum_{i=1}^{m}\sigma_i a_i
\]
\end{definition}

\begin{definition}
Let $A \subset [0,1]^m$ and $\sigma \in \{-1,1\}^m$. 
The Komlos complexity of $A$ is 
\[
K_m(A) =  \frac{1}{m}\min_{\sigma} \sup_{a \in A}  \sum_{i=1}^{m}\sigma_i a_i
\]
\end{definition}
\noindent It well known that $R_n$ is a good measure for generalization when data is chosen uniformly at random (or drawn from an unknown distribution). See \cite{Bartlett:2003:RGC:944919.944944} for other notions complexity and relationships between them. We claim that $K_n$ is the measure more suitable for active learning, for creating coresets, and for creating mergeable sketches.

The name Komlos Complexity is due to the Komlos Conjecture. It claims that $\min_\sigma \max_f \|\sum_i \sigma_i f_i\|_\infty = O(1)$ for unit vectors $f$. The correctness of the Komlos Conjecture is still a fundamental open problem in the study of discrepancy theory.  The definition is useful, however, regardless of the correctness or provability of the conjecture.  
It is obvious that  $K_m(\F) \le R_m(\F)$ for any family $\F$. We show later that $K_m(\F) = o(R_m(\F))$ for a wide range of interesting problems in machine learning.

%\begin{theorem}
%Let $F(a) = \sum_i a_i$ for $a \in A$ and let $K_m(A) = O(c/m^\gamma)$ for any constant $\gamma \in (0,1]$.
%There exist a coreset for $A$ such that $\tilde{F}(a) = \sum_{i} w_i a_i$, $\operatorname{supp}(w) \le m$ and 
%$$
%\forall \; a \in A \;\;\;  |F(a) - \tilde F(a)| \le \eps n
%$$
%\end{theorem}
%\begin{proof}
%By halving the points repeatedly \el{TODO}.
%\end{proof}

@@

\noindent Proving both facts is trivial. 
At step one, create a coreset of size at most $n/2$ of items of weight $2$ and incur an error of at most $c$ or $c \sqrt{n}$.
Then, we repeat the process and create a coreset of size $n/4$ of items of weight $4$. Here, you incur error of $2c$ of $2\sqrt{n/2}  = \sqrt{2n}$.
Note that the sum of errors is asymptotically dominated by last iterations. 
Halting the compression at $\Theta(1/\eps)$ or $\Theta(1/\eps^2)$ items respectively achieves the goal.

@@

\section{General Framework for Sketching}

The high level idea is the following. 
Consider a stream of the data points $x_i$. 
We maintain a buffer that can hold up to $m$ points.
Once collecting $m$ data points it performs a \emph{compact} operation that outputs at most $m/2$ data points of weight $2$ representing the original $m$ inputs. The compact operation partitions the $m$ points into two sets whose corresponding sums are very close. 
This is made formal with the following definitions: 
Let us consider a signed sum of error function $E(q) = \sum_{i=1}^{m} \sigma_i f(x_i,q)$ where $\sigma_i \in \{-1,1\}$.
Now consider $F_1(q) = F(q) + E(q) = \sum_{i=1}^{m} f(x_i,q)  + \sum_{i=1}^{m} \sigma_i f(x_i,q)  = 2 \sum_{i | \sigma_i=1} f(x_i,q)$ and the corresponding $F_{-1}$. We have that both $F_1(q), F_{-1}(q)$ are approximations for $F(q)$ obtained by coresets of item weights of $2$ and an error at most $|F_1(q) - F(q)| \le  |E(q)|$.

%
%We begin with describing the basic operation of compaction and the notion of flat error functions. 
%
%Let us consider a signed sum of error function $E(q) = \sum_{i=1}^{n} \sigma_i f(x_i,q)$.
%Now consider $\tilde F(q) = F(q) + E(q) = \sum_{i=1}^{n} f(x_i,q)  + \sum_{i=1}^{n} \sigma_i f(x_i,q)  = 2 \sum_{i | \sigma_i=1} f(x_i,q)$.
%Therefore $\tilde F(q)$ is a coreset for $F(q)$ with item weights of $2$ and error at most $|\tilde F(q) - F(q)| \le \max_q | E(q)|$.
Note that the above is true for any choice of signs $\sigma_i$, specifically, for those minimizing $\max_q | E(q)|$.
Note that we can select signs such that $|E(q)| \le m K_m(\F)$.
We claim that a function is sketch-able if $K_m(\F) = O(m^\gamma)$ for any constant $\gamma \in [0,1)$.

In what follows, we prove the following.
\begin{theorem} \label{thm:streaming}
For any function $f$ with a corresponding cthere exists an unbiased fully-mergeable streaming coreset algorithms of size 
$O\left(\log^2\left(\min(n, \log(1/\delta)) \right)/\eps\right)$ whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require knowledge of the stream length $n$.
\end{theorem}

\begin{theorem}
For any function $f$ with a corresponding omlos Complexity of $K_m = O(1/\sqrt{m})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
$O\left(\log^3\left(\min(n, \log(1/\delta)) \right)/\eps^2\right)$ whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require knowledge of the stream length $n$.
\end{theorem}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
%$O(\log^2(\log(1/\delta))/\eps^2)$ whose error is at most $\eps n$ and failure probability at most $\delta$.
%\end{fact}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(1)$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps)$ whose error is at most $\eps n$.
%\end{fact}
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps^2)$ whose error is at most $\eps n$.
%\end{fact}

\begin{lemma} \label{lem:compactor}
Let $k$ be an integer and assume we have black box access to a solver that for any $m$ inputs $x_1,\ldots,x_m$ obtains $m$ signs $\sigma_1,\ldots,\sigma_m$ such that
$$\max_q \left| \sum_i \sigma_i f(x_i, q)\right| \leq K_m$$
\el{I need to think about this absolute value here. Is it needed?}
Then there exist a streaming algorithm requiring a memory buffer of $k$ input items that given a stream of length $n$ outputs a stream $z_1,\ldots,z_m$ with the following properties
\begin{itemize}
\item $\{z_i\}_i$ is a subset of $\{x_i\}$
\item $\E[m] = n/2$
\item For any fixed query $q$, the error associated with $q$, defined as
$$\text{Err}(q) = 2\sum_{i=1}^m f(z_i,q) - \sum_{i=1}^n f(x_i,q)  $$
Can be decomposed as a sum
$$\text{Err}(q) = \sum_{j=1}^{n/k} \text{Err}_j(q)$$
where the different $\text{Err}_j(q)$ are independent, have mean 0 and $|\text{Err}_j(q)| \leq \rho k$
\end{itemize}
The last property ensures that deterministically $|\text{Err}(q)| \leq \rho n$, and that w.h.p.\ (via Chernoff's inequality) for a fixed $q$ 
$$|\text{Err}(q)| \lesssim \rho k\sqrt{n/k} = \rho \sqrt{kn}$$
\end{lemma}
\begin{proof}
The algorithm operates as follows. It keeps a buffer of $k$ items. Once the buffer fills with $x_1,\ldots,x_k$ it obtains the signs guaranteeing
$$\max_q \left| \sum_i \sigma_i f(x_i, q)\right| \leq \rho k$$
Then, the algorithm output to the stream the with probability $1/2$ each data points $\{x_i \; | \; \sigma_i = 1\}$ or $\{x_i \; | \; \sigma_i = -1\}$ with twice the weight.
$$\sum_{i ,\; \sigma_i=1} 2f(x_i, q) = \sum_{i} f(x_i, q) +  \sum_{i} \sigma_i f(x_i, q)$$
$$\sum_{i ,\; \sigma_i=-1} 2f(x_i, q) = \sum_{i} f(x_i, q) - \sum_{i} \sigma_i f(x_i, q)$$

>>>>

\section{Komlos Complexity}
We begin by formally defining the Komols complexity and by giving some general results about for problems with low Komols complexity. The Komols complexity of a sequence set is defined similarly to it Rademacher complexity.
%\begin{definition}
%Let $A \subset [0,1]^m$ and $\sigma \in \{-1,1\}^m$ uniformly. 
%The Rademacher complexity of $\F$ is 
%\[
%R_m(\F) =  \frac{1}{m}\E_{\sigma} \sup_{a \in A}  \sum_{i=1}^{m}\sigma_i a_i
%\]
%\end{definition}
%
%\begin{definition}
%Let $A \subset [0,1]^m$ and $\sigma \in \{-1,1\}^m$. 
%The Komlos complexity of $A$ is 
%\[
%K_m(A) =  \frac{1}{m}\min_{\sigma} \sup_{a \in A}  \left|\sum_{i=1}^{m}\sigma_i a_i\right|
%\]
%\end{definition}

\begin{definition}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$ uniformly. 
The Rademacher complexity of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
R_m(\F) =  \frac{1}{m}\E_{\sigma} \sup_{f \in \F}  \sum_{i=1}^{m}\sigma_i f(x_i)
\]
\end{definition}
%
\begin{definition}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The Komlos complexity of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(\F) =  \frac{1}{m}\min_{\sigma} \sup_{f \in \F}  \left|\sum_{i=1}^{m}\sigma_i f(x_i)\right|
\]
The Komlos Complexity $K_m(\F)$ without a reference set $\{x_1,\ldots,x_m\}$ is the upper bound on any such set.
\end{definition}

\noindent It well known that $R_m$ is a good measure for generalization when data is chosen uniformly at random (or drawn from an unknown distribution). See \cite{Bartlett:2003:RGC:944919.944944} for other notions of complexity and relationships between them. We claim that $K_m$ is the measure more suitable for active learning, for creating coresets, and for creating mergeable sketches.

\el{Add a much longer explanation about the connection to discrepancy and the ideas behind this construction} The name Komlos Complexity is due to the Komlos Conjecture. It claims that $\min_\sigma \max_f \|\sum_i \sigma_i f_i\|_\infty = O(1)$ for unit vectors $f$. The correctness of the Komlos Conjecture is still a fundamental open problem in the study of discrepancy theory.  The definition is useful, however, regardless of the correctness or provability of the conjecture.  
It is obvious that  $K_m(\F) \le R_m(\F)$ for any family $\F$. We show later that $K_m(\F) = o(R_m(\F))$ for a wide range of interesting problems in machine learning.

%\begin{theorem}
%Let $F(a) = \sum_i a_i$ for $a \in A$ and let $K_m(A) = O(c/m^\gamma)$ for any constant $\gamma \in (0,1]$.
%There exist a coreset for $A$ such that $\tilde{F}(a) = \sum_{i} w_i a_i$, $\operatorname{supp}(w) \le m$ and 
%$$
%\forall \; a \in A \;\;\;  |F(a) - \tilde F(a)| \le \eps n
%$$
%\end{theorem}
%\begin{proof}
%By halving the points repeatedly \el{TODO}.
%\end{proof}
\subsection{General Framework for Coresets}
Consider the function $F = \sum_{i=1}^{m} f(x_i,q)$ and the signed sum of error function $E(q) = \sum_{i=1}^{m} \sigma_i f(x_i,q)$ where $\sigma_i \in \{-1,1\}$.
Now consider $F_1(q) = F(q) + E(q) = \sum_{i=1}^{m} f(x_i,q)  + \sum_{i=1}^{m} \sigma_i f(x_i,q)  = 2 \sum_{i | \sigma_i=1} f(x_i,q)$ and similarly $F_{-1}$. We have that both $F_1(q), F_{-1}(q)$ are approximations for $F(q)$ obtained by coresets of item weights of $2$ and an error at most $|F_1(q) - F(q)| =  |E(q)|$.
%
The above is true for any choice of signs $\sigma_i$, specifically, for those minimizing $\max_q | E(q)|$.
Note that we can select signs such that $|E(q)| \le m K_m(\F)$.
A function is sketch-able if $K_m(\F) = O(m^\gamma)$ for any constant $\gamma \in [0,1)$.
Typically we see that $K_m(\F) = O(c/m)$ or $K_m(\F)= c/\sqrt{m}$ for some $c$ which does not depend on $m$.

@@

\noindent Proving both facts is trivial. 
At step one, create a coreset of size at most $n/2$ of items of weight $2$ and incur an error of at most $c$ or $c \sqrt{n}$.
Then, we repeat the process and create a coreset of size $n/4$ of items of weight $4$. Here, you incur error of $2c$ or $2c\sqrt{n/2}  = \sqrt{2n}$.
Note that the sum of errors is asymptotically dominated by last iterations. 
Halting the compression at $\Theta(c/\eps)$ or $\Theta(c/\eps^2)$ items respectively achieves the goal.

@@

\el{I believe we that the correctness of the Komlos conjecture implies equivalence between coresets limited to the input and general coresets. Specifically, a constructive positive resolution will give an algorithm for converting any coresets to signs $\sigma$ and select inputs from a sequence.}

\subsection{General Framework for Sketching}

The high level idea is the following. 
Consider a stream of the data points $x_i$. 
We maintain a buffer that can hold up to $m$ points.
Once collecting $m$ data points it performs a \emph{compact} operation that outputs at most $m/2$ data points of weight $2$ representing the original $m$ inputs. The compact operation partitions the $m$ points into two sets whose corresponding sums are very close. 
This is made formal with the following definitions:

 %
In what follows, we prove the following.
\begin{theorem} \label{thm:streaming}
For any function $f$ with a corresponding Komlos Complexity $K_m = O(c/m)$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
$O\left(c\log^2\left(\min(n, \log(1/\delta)) \right)/\eps\right)$ whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require a priori knowledge of the stream length $n$.
\end{theorem}

\begin{theorem}
For any function $f$ with a corresponding Komlos Complexity of $K_m = O(c/\sqrt{m})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
$O\left(c\log^3\left(\min(n, \log(1/\delta)) \right)/\eps^2\right)$ whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require a priori knowledge of the stream length $n$.
\end{theorem}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
%$O(\log^2(\log(1/\delta))/\eps^2)$ whose error is at most $\eps n$ and failure probability at most $\delta$.
%\end{fact}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(1)$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps)$ whose error is at most $\eps n$.
%\end{fact}
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps^2)$ whose error is at most $\eps n$.
%\end{fact}

\begin{lemma} \label{lem:compactor}
Let $m$ be an integer and assume we have black box access to a solver that for any $m$ inputs $x_1,\ldots,x_m$ obtains $m$ signs $\sigma_1,\ldots,\sigma_m$ such that
$$\max_q \left| \sum_{i=1}^{m} \sigma_i f(x_i, q)\right| \in O(K_m)$$
Note that such signs are guarantied to exist by the definition of the Komols Complexity. 
In fact, the quantity above is likely much smaller than $K_m$ because it is computed for a specific set of points $x_i$ and the worst such possible set.

Then there exist a streaming algorithm requiring a memory buffer of $k$ input items that given a stream of length $n$ outputs a stream $z_1,\ldots,z_m$ with the following properties
\begin{itemize}
\item $\{z_i\}_i$ is a subset of $\{x_i\}$
\item $\E[m] = n/2$
\item For any fixed query $q$, the error associated with $q$, defined as
$$\text{Err}(q) = 2\sum_{i=1}^m f(z_i,q) - \sum_{i=1}^n f(x_i,q)  $$
Can be decomposed as a sum
$$\text{Err}(q) = \sum_{j=1}^{n/k} \text{Err}_j(q)$$
where the different $\text{Err}_j(q)$ are independent, have mean 0 and $|\text{Err}_j(q)| \leq \rho k$
\end{itemize}
The last property ensures that deterministically $|\text{Err}(q)| \leq \rho n$, and that w.h.p.\ (via Chernoff's inequality) for a fixed $q$ 
$$|\text{Err}(q)| \lesssim \rho k\sqrt{n/k} = \rho \sqrt{kn}$$
\end{lemma}
\begin{proof}
The algorithm operates as follows. It keeps a buffer of $k$ items. Once the buffer fills with $x_1,\ldots,x_k$ it obtains the signs guaranteeing
$$\max_q \left| \sum_i \sigma_i f(x_i, q)\right| \leq \rho k$$
Then, the algorithm output to the stream the with probability $1/2$ each data points $\{x_i \; | \; \sigma_i = 1\}$ or $\{x_i \; | \; \sigma_i = -1\}$ with twice the weight.
$$\sum_{i ,\; \sigma_i=1} 2f(x_i, q) = \sum_{i} f(x_i, q) +  \sum_{i} \sigma_i f(x_i, q)$$
$$\sum_{i ,\; \sigma_i=-1} 2f(x_i, q) = \sum_{i} f(x_i, q) - \sum_{i} \sigma_i f(x_i, q)$$
